---
title: "Preliminary on Credit card data"
author: "Sunny Jing"
date: "2/6/2017"
output:
  html_document: default
  pdf_document: default
---

```{r}
datapath<-"/Users/dongpingjing/Dropbox/Data Mining/Project Folder/Revised Raw Data Set"
original<-read.csv(file=paste(datapath,"LoanStats3b.csv",sep="/"))

# Kepp only 36 month term data:
original<-original[original$term==" 36 months",]
original$term<-NULL

loan.data<-original

# Response varible:
loan.data$defaulted[loan.data$loan_status=="Fully Paid"]<-0
loan.data$defaulted[is.na(loan.data$defaulted)]<-1
loan.data$defaulted<-factor(loan.data$defaulted)
loan.data$loan_status<-NULL

# Remove text value variables:
loan.data$url<-NULL
loan.data$emp_title<-NULL
loan.data$desc<-NULL
loan.data$zip_code<-NULL
loan.data$addr_state<-NULL
loan.data$title<-NULL  # too messy

# Remove data variables and ID:
loan.data$id<-NULL
loan.data$member_id<-NULL
loan.data$issue_d<-NULL
loan.data$earliest_cr_line<-NULL
loan.data$last_pymnt_d<-NULL
loan.data$next_pymnt_d<-NULL #Some fully paid account still have scheduled next payment in Feb 2017.
loan.data$last_credit_pull_d<-NULL
loan.data$grade<-NULL
loan.data$sub_grade<-NULL

# variables to remove because of blank:
blank<-apply(loan.data,2,function(x){sum(is.na(x))})
loan.data[,blank==nrow(loan.data)]<-NULL

# Make sure class right:
loan.data$policy_code<-factor(loan.data$policy_code)
loan.data$revol_util<-as.numeric(levels(loan.data$revol_util)[loan.data$revol_util])

# variables to remove because of unified value:
one.value<-apply(loan.data,2,function(x){length(unique(x))})
loan.data[,one.value==1]<-NULL

# Remove variables regarding current credit card infomation: 
# ("loan_amnt", "funded_amnt", funded_amnt_inv", "int_rate", "installment","out_prncp","out_prncp_inv","total_pymnt""total_pymnt_inv","total_rec_prncp","total_rec_int","total_rec_late_fee","recoveries","collection_recovery_fee""last_pymnt_amnt"")
loan.data[,c(1:5,22:31)]<-NULL
```


## Finding and imputing for missing values:
```{r}
missing<-apply(loan.data,2,function(x){sum(is.na(x))})
colnames.missing<-colnames(loan.data[,missing!=0])
colnames.missing<-colnames.missing[order(colnames.missing)]
colnames.missing

# fill in variables that represent number of months with the 99 quantile of the data:
colname<-colnames.missing[c(4:14,27)]
for (i in colname){         
  loan.data[ which(is.na(loan.data[,i])),i]<-quantile(loan.data[which(!is.na(loan.data[,i])),i], 0.99)
}

#fill in variables that represent amounts with 0:
colname<-colnames.missing[c(1:3,14:26,28:34)]
for (i in colname){
  loan.data[ which(is.na(loan.data[,i])),i] <-0
}
```


## Select the varaibles for logistic model:

__categorical variables__
```{r}
cat<-NULL
n=ncol(loan.data)
for (i in 1:n){
  cat[i]<-class(loan.data[,i])
}
cat.var<-loan.data[,cat=="factor"]

# Try Tree regression:
library("caret")
library("rpart")
tree<-rpart(defaulted~.,data=cat.var,control=rpart.control(minsplit=30L,cp=0,xval=10L,maxsurrogate = 0))
tree$cptable
rpart.plot(tree,extra=2)

# Conduct step wise analysis:
logistic.cat<-glm(defaulted~.,data=cat.var,family=binomial(link=logit))
summary(logistic.cat)
stepwise.cat<-step(logistic.cat,)
summary(stepwise.cat)

# Recategorize purpose:
levels(loan.data$purpose)[c(1:8,12,13)]<-"consolidated"

# Recategorize employment length:
levels(loan.data$emp_length)[c(1:11)]<-"1 year +"

```

__numeric variables__
```{r}
# Checking correlation between numeric variables:
num.var<-loan.data[,cat!="factor"]
correlation<-round(cor(num.var),3)

logistic.full<-glm(defaulted~.,data=loan.data,family=binomial(link=logit))
summary(logistic.full)
logistic.full$coefficients
# Remove other insignificant variables
to.remove<-names(summary(logistic.full)$coefficients[,1][round(summary(logistic.full)$coefficients,2)[,1]==0 & summary(logistic.full)$coefficients[,4]>=0.01])
to.remove
model.data<-loan.data[,-which(colnames(loan.data) %in% to.remove)]
colnames(model.data)
```


Train and test:
```{r}
# Take 63.2% for training and the rest for testing:
n=nrow(model.data)
set.seed(2345341)
rownumbers<-sample(seq_len(nrow(model.data)),size=n*0.632)
train<-model.data[rownumbers,]
test<-model.data[-rownumbers,]
```


Logistic Regression:
```{r}
log.model.full<-glm(defaulted~.,data=train,family=binomial(link=logit))
summary(log.model.full)
stepwise<-step(log.model.full)

lda<-lda(formula,train)


formula<-defaulted ~ emp_length + home_ownership + annual_inc + verification_status + purpose + dti + delinq_2yrs + inq_last_6mths +  open_acc + revol_util + total_acc + acc_open_past_24mths +  bc_open_to_buy + bc_util + mo_sin_old_rev_tl_op + mths_since_recent_bc +  mths_since_recent_inq + num_actv_bc_tl + num_bc_sats + num_il_tl +  num_rev_tl_bal_gt_0 + num_sats + num_tl_30dpd + num_tl_90g_dpd_24m +  num_tl_op_past_12m + percent_bc_gt_75 + pub_rec_bankruptcies

log.model<-glm(formula,data=train,family=binomial(link=logit))
summary(log.model)
anova(log.model,log.model.full,test="Chisq")

learn<-glm(defaulted~percent_bc_gt_75,train,family=binomial(link=logit))
anova(log.model.full,learn,test="Chisq")

# Try decision tree model:
tree.model<-rpart(formula,data=train,control=rpart.control(minsplit=30L,cp=0,xval=10L,maxsurrogate = 0))
tree.model1<-rpart(formula,data=train,control=rpart.control(minsplit=30L,cp=2.336995e-04,xval=10L,maxsurrogate = 0))
tree.model1$cptable
rpart.plot(tree.model1)
```

## VIF:

* Variable Group 1: Number of Accounts,
open_acc, total_acc, mort_acc, num_actv_bc_tl, num_actv_rev_tl, num_bc_sats, num_bc_tl, num_il_tl, num_op_rev_tl, num_rev_accts,num_rev_tl_bal_gt_0, num_sats.

* Variable Group 2: Age (= length of credit history),
earliest_cr_line, mo_sin_old_il_acct, mo_sin_old_rev_tl_op,

* Variable Group 3: Balance,
revol_bal, tot_cur_bal, avg_cur_bal, total_bal_ex_mort,

* Variable Group 4: Derogatory (~ bad) records;
delinq_2yrs, mths_since_last_delinq, mths_since_last_record, pub_rec, collections_12_mths_ex_med, mths_since_last_major_derog, acc_now_delinq, tot_coll_amt, chargeoff_within_12_mths, delinq_amnt, mths_since_recent_bc_dlq, mths_since_recent_revol_delinq, num_accts_ever_120_pd, num_tl_120dpd_2m, num_tl_30dpd, num_tl_90g_dpd_24m, num_tl_op_past_12m, pct_tl_nvr_dlq, pub_rec_bankruptcies, tax_liens.

* Variable Group 5: Inquiries;
inq_last_6mths,mths_since_recent_inq.

* Variable Group 6: Number of Credit Lines/Limits;
total_rev_hi_lim, tot_hi_cred_lim, total_bc_limit, total_il_high_credit_limit.

* Variable Group 7: New (=Recently Opened) Accounts;
acc_open_past_24mths, mo_sin_rcnt_rev_tl_op, mo_sin_rcnt_tl,mths_since_recent_bc.

* Variable Group 8: Utilization;
revol_util, bc_util, bc_open_to_buy,percent_bc_gt_75.

* Variable Group 9: Internal Variables;
* They appear to be internal variables not available at application but created later to include future information relative to application.;
out_prncp, out_prncp_inv, policy_code, recoveries, collection_recovery_fee.

```{r}
library("car")
vif(log.model)

# Drop num_sats in Group 1:
formula<-defaulted ~ emp_length + home_ownership + annual_inc + verification_status + purpose + dti + delinq_2yrs + inq_last_6mths +  open_acc + revol_util + total_acc + acc_open_past_24mths +  bc_open_to_buy + bc_util + mo_sin_old_rev_tl_op + mths_since_recent_bc +  mths_since_recent_inq + num_actv_bc_tl + num_bc_sats + num_il_tl +  num_rev_tl_bal_gt_0 + num_tl_30dpd + num_tl_90g_dpd_24m +  num_tl_op_past_12m + percent_bc_gt_75 + pub_rec_bankruptcies

log.model.vif<-glm(formula,data=train,family=binomial(link=logit))
vif(log.model.vif)

# Drop num_actv_bc_tl in Group1:
formula<-defaulted ~ emp_length + home_ownership + annual_inc + verification_status + purpose + dti + delinq_2yrs + inq_last_6mths +  open_acc + revol_util + total_acc + acc_open_past_24mths +  bc_open_to_buy + bc_util + mo_sin_old_rev_tl_op + mths_since_recent_bc +  mths_since_recent_inq + num_bc_sats + num_il_tl +  num_rev_tl_bal_gt_0 + num_tl_30dpd + num_tl_90g_dpd_24m +  num_tl_op_past_12m + percent_bc_gt_75 + pub_rec_bankruptcies

log.model.vif<-glm(formula,data=train,family=binomial(link=logit))
vif(log.model.vif)

# Drop bc_util in Group 8:
formula<-defaulted ~ emp_length + home_ownership + annual_inc + verification_status + purpose + dti + delinq_2yrs + inq_last_6mths +  open_acc + revol_util + total_acc + acc_open_past_24mths +  bc_open_to_buy + mo_sin_old_rev_tl_op + mths_since_recent_bc +  mths_since_recent_inq + num_bc_sats + num_il_tl +  num_rev_tl_bal_gt_0 + num_tl_30dpd + num_tl_90g_dpd_24m +  num_tl_op_past_12m + percent_bc_gt_75 + pub_rec_bankruptcies

log.model.vif<-glm(formula,data=train,family=binomial(link=logit))
vif(log.model.vif)

# Drop total_acc in Group 1:
formula<-defaulted ~ emp_length + home_ownership + annual_inc + verification_status + purpose + dti + delinq_2yrs + inq_last_6mths +  open_acc + revol_util + acc_open_past_24mths +  bc_open_to_buy + mo_sin_old_rev_tl_op + mths_since_recent_bc +  mths_since_recent_inq + num_bc_sats + num_il_tl +  num_rev_tl_bal_gt_0 + num_tl_30dpd + num_tl_90g_dpd_24m +  num_tl_op_past_12m + percent_bc_gt_75 + pub_rec_bankruptcies

log.model.vif<-glm(formula,data=train,family=binomial(link=logit))
vif(log.model.vif)

# Drop open_acc in Group 1:
formula<-defaulted ~ emp_length + home_ownership + annual_inc + verification_status + purpose + dti + delinq_2yrs + inq_last_6mths + revol_util + acc_open_past_24mths +  bc_open_to_buy + mo_sin_old_rev_tl_op + mths_since_recent_bc +  mths_since_recent_inq + num_bc_sats + num_il_tl +  num_rev_tl_bal_gt_0 + num_tl_30dpd + num_tl_90g_dpd_24m +  num_tl_op_past_12m + percent_bc_gt_75 + pub_rec_bankruptcies

log.model.vif<-glm(formula,data=train,family=binomial(link=logit))
vif(log.model.vif)

# Drop revol_util in Group 8:
formula<-defaulted ~ emp_length + home_ownership + annual_inc + verification_status + purpose + dti + delinq_2yrs + inq_last_6mths + acc_open_past_24mths +  bc_open_to_buy + mo_sin_old_rev_tl_op + mths_since_recent_bc +  mths_since_recent_inq + num_bc_sats + num_il_tl +  num_rev_tl_bal_gt_0 + num_tl_30dpd + num_tl_90g_dpd_24m +  num_tl_op_past_12m + percent_bc_gt_75 + pub_rec_bankruptcies

log.model.vif<-glm(formula,data=train,family=binomial(link=logit))
vif(log.model.vif)

# Drop num_bc_sats in Group 1:
formula<-defaulted ~ emp_length + home_ownership + annual_inc + verification_status + purpose + dti + delinq_2yrs + inq_last_6mths + acc_open_past_24mths +  bc_open_to_buy + mo_sin_old_rev_tl_op + mths_since_recent_bc +  mths_since_recent_inq + num_il_tl +  num_rev_tl_bal_gt_0 + num_tl_30dpd + num_tl_90g_dpd_24m +  num_tl_op_past_12m + percent_bc_gt_75 + pub_rec_bankruptcies

log.model.vif<-glm(formula,data=train,family=binomial(link=logit))
vif(log.model.vif)

# Drop acc_open_past_24mths in Group 7:
formula<-defaulted ~ emp_length + home_ownership + annual_inc + verification_status + purpose + dti + delinq_2yrs + inq_last_6mths  +  bc_open_to_buy + mo_sin_old_rev_tl_op + mths_since_recent_bc +  mths_since_recent_inq + num_il_tl +  num_rev_tl_bal_gt_0 + num_tl_30dpd + num_tl_90g_dpd_24m +  num_tl_op_past_12m + percent_bc_gt_75 + pub_rec_bankruptcies

log.model.vif<-glm(formula,data=train,family=binomial(link=logit))
vif(log.model.vif)

# Drop mths_since_recent_inq in Group 5:
formula<-defaulted ~ emp_length + home_ownership + annual_inc + verification_status + purpose + dti + delinq_2yrs + inq_last_6mths  +  bc_open_to_buy + mo_sin_old_rev_tl_op + mths_since_recent_bc + num_il_tl +  num_rev_tl_bal_gt_0 + num_tl_30dpd + num_tl_90g_dpd_24m +  num_tl_op_past_12m + percent_bc_gt_75 + pub_rec_bankruptcies

library("MASS")
lda.model<-lda(formula,train)

log.model.vif<-glm(formula,data=train,family=binomial(link=logit))
vif(log.model.vif)

final.model<-log.model.vif
```

## Boostrapping:

```{r}
bootstrapping <-data.frame(matrix(rep(0,22000),nrow=1000,ncol=24))
colnames(bootstrapping)<-rownames(summary(final.model)$coefficients)

for (i in 1:1000){
  set.seed(12345*i)
  rnumber<-sample(seq_len(nrow(model.data)),size=n*0.632)
  boots.train<-model.data[rnumber,]
  log.model<-glm(formula,data=boots.train,family=binomial(link=logit))
  bootstrapping [i,]<-summary(log.model)$coefficients[,1]
}

boots.coeff<-apply(bootstrapping ,2,mean)
summary(final.model)
final.model$coefficients <- boots.coeff
```

## Test
```{r}
pred.test<-predict.glm(final.model,test,type="response")
prediction.test<-ifelse(pred.test>0.5,1,0)

confusionMatrix(prediction.test,test$defaulted)

lift.chart<-as.data.frame(cbind(pred=pred.test[order(pred.test,decreasing=TRUE)],defaulted=test$defaulted[order(pred.test,decreasing=TRUE)]))
lift.chart[,2]<-lift.chart[,2]-1
head(lift.chart,10)

n=round(nrow(lift.chart)/10)
for (i in 1:10){
  cumulative.lift.chart$cumul.lift[i]<-sum(lift.chart[(n*(i-1)+1):(n*i),2]/n)
}
overall<-sum(lift.chart[1:(n*10),2])/(n*10)

# Lift Chart:
library("ggplot2")
barplot(cumulative.lift.chart[,2],ylim=c(0,0.5),)
abline(h=overall,col="red")
```

## Assign scores:
```{r}
final.pred<-predict(final.model,model.data,type="response")
scores<-round(850-(850-300)*final.pred)
```

## Decide a cutoff score based on percentage of number of people:
```{r}
deciles<-quantile(scores, probs = seq(0,1, 0.10))
deciles

Decile.analysis<-data.frame(Percentile=c(1:10),
                            N=rep(round(nrow(model.data)/10),10),
                            Probability=round(rev(tapply(final.pred,findInterval(final.pred,quantile(final.pred, probs =seq(0,1, 0.10)),all.inside = TRUE),mean)),2),
                            Min_Score=tapply(scores,findInterval(scores,deciles,all.inside = TRUE),min),
                            Median_Score=tapply(scores,findInterval(scores,deciles,all.inside = TRUE),median),
                            Max_Score=tapply(scores,findInterval(scores,deciles,all.inside = TRUE),max))
Decile.analysis$"E(Num_Default)"<-round(Decile.analysis$N*Decile.analysis$Probability)
for (i in 1:10){
  Decile.analysis$"Cum_E(Num_Default)"[i]<-sum(Decile.analysis$"E(Num_Default)"[i:10])
}

Decile.analysis$Percentage_Default<-round(Decile.analysis$`Cum_E(Num_Default)`/nrow(model.data),2)
Decile.analysis

```

## Decide a cutoff score based on a fixed money amount:

```{r}
ordered.loan.amnt<-original$loan_amnt[order(final.pred,decreasing = TRUE)]
k=round(length(ordered.loan.amnt)/10)

Decile.analysis$Ave_Amount<-tapply(ordered.loan.amnt,c(rep(1,k),rep(2,k),rep(3,k),rep(4,k),rep(5,k),rep(6,k),rep(7,k),rep(8,k),rep(9,k),rep                             (10,k+1)),mean)
Decile.analysis$"E(default_amnt)"<-Decile.analysis$`E(Num_Default)`*Decile.analysis$Ave_Amount
for (i in 1:10){
  Decile.analysis$"Cum_E(default_amnt)"[i]<-sum(Decile.analysis$`E(default_amnt)`[i:10])
}

Decile.analysis

write.csv(Decile.analysis,"Decile.analysis.csv")

plot(final.pred,scores)
```

## KS Stats:
```{r}
# KS statistics:
library("KScorrect")
kstest<-ks.test(scores[scores<755],scores[scores>=755])
kstest
```